A Formalisation of Sturm's Theorem\\in Isabelle/HOL

### 2013-12-17 slides/slides.tex
Updated slides, tuned polynomials

\begin{frame}{Sturm sequence}
But what is a Sturm sequence of $P$?\\
A (reasonably) general definition:\\
\begin{itemize}
\item $P_0 = P$
\item $P_0$ and $P_1$ have no common roots
\item $P_{n-1}$ (last element) does not change its sign
\item if $x_0$ is root of $P_0$: $P_0P_1(x)<0$ in some left-NH and $P_0P_1(x)>0$ in some right-NH of $x_0$
\item if $x_0$ is root of another $P_i$: $P_{i-1}P_{i+1}(x_0) < 0$
\end{itemize}
\end{frame}

@@

\begin{frame}{Proving Sturm's Theorem}
Assume we already have a Sturm chain. Why does it count roots?\\
Follow $x \mapsto \sigma(P_0, \ldots, P_{n-1}; x)$ passing over $\RR$. Obviously, it can only change at $x_0$ if one of the $P_i$ has a root at $x_0$
\begin{itemize}
\item if $P_i \neq P_0$ has $x_0$ as a root, $P_{i-1}P_{i+1}(x_0) < 0$\\
$\Rightarrow$ signs of $P_{i-1}$, $P_{i+1}$ are $\neq 0$, opposite, constant in NH of $x_0$\\
$\Rightarrow$ signs $[1,\, \_\,, -1]$ or $[-1,\, \_\,, 1]$ in the entire NH, i.e. one sign change\\
$\Rightarrow$ total number of sign changes not influenced\\[1mm]
\item if $P_0$ has $x_0$ as root, $P_0P_1(x_0) < 0$ in left-NH of $x_0$, $>0$ in right-NH\\
$\Rightarrow$ signs are different left of $x_0$ and the same right of $x_0$\\
$\Rightarrow$ total number of sign changes decreases by one
\end{itemize}
\end{frame}

\begin{frame}{Proving Sturm's Theorem}
Formal proof: a lot of induction on the sequences and number of roots\\[2mm]
$\Longrightarrow$ messy and not terribly interesting
\end{frame}

@@

\begin{frame}{Construction Sturm sequences}
Why does it work? Nonobvious parts:\\[3mm]

If $x_0$ is root of $P_0 = P$: $PP'(x_0) < 0$ in left-NH and $PP'(x_0) > 0$ in right-NH\\
-- pick neighbourhood without roots of $P_0$ and $P_1$ (except for $x_0$),\\
\hskip3mm apply mean value theorem\\[3mm]

@@

\begin{frame}{Construction Sturm sequences}
In case of multiple roots: Let $D := \mathrm{gcd}(P,P')$. Then:
\begin{description}
\item[The obvious way:]\ \\
\begin{itemize}
\item compute canonical Sturm chain of $P / D$\\ (\enquote{divide out} multiple roots)
\end{itemize}
\item[The clever way:]\ \\
\begin{itemize}
\item we can compute the canonical Sturm chain of $P$ and divide by $D$ afterwards
\item but: if $D(x)\neq 0$, dividing by $D$ does not change the number of sign changes at $x$
\item $\Longrightarrow$ unless the interval bounds are multiple roots, we can use the canonical construction without changes
\end{itemize}
\end{description}
\end{frame}

\begin{frame}{Making a decision procedure}
\textit{count\_roots\_between\ p\ a\ b}: picks the most efficient Sturm chain construction and:
$$\textit{count\_roots\_between}\ p\ a\ b\ \ =\ \ |\{x.\ a < x\ \wedge\ x \leq b\ \wedge\ p(x) = 0\}|$$\vskip2mm

@@

\begin{frame}{Making a decision procedure}
Some more fluff:
\begin{itemize}
\item and/or: count $x$ with $$P(x)=0\ \wedge\ Q(x)=0\hskip5mm\mathrm{or}\hskip5mm P(x)=0\ \vee\ Q(x)=0$$
\item $\forall$-inequalities: $$\forall x.\ P(x) \neq Q(x)\ \wedge\ R(x) \neq S(x)\ \vee\ T(x)\neq U(x)$$
\item $\forall$ with $<$ and $>$: $$\forall x.\ P(x) < Q(x)\ \wedge\ R(x) > S(x)\ \vee\ T(x)\neq U(x)$$
\end{itemize}
\end{frame}

>>>>

\begin{frame}{Sturm sequence}
But what is a Sturm sequence of $P$?\\
A (reasonably) general definition:\\
\begin{itemize}
\item $P_0 = P$\pause
\item $P_0$ and $P_1$ have no common roots\pause
\item $P_{n-1}$ (last element) does not change its sign\pause
\item if $x_0$ is root of $P_0$: $P_0P_1(x)<0$ in some left-NH and $P_0P_1(x)>0$ in some right-NH of $x_0$\pause
\item if $x_0$ is root of another $P_i$: $P_{i-1}P_{i+1}(x_0) < 0$
\end{itemize}
\end{frame}

@@

\begin{frame}{Proving Sturm's Theorem}
Assume we already have a Sturm chain. Why does it count roots?\\
Follow $x \mapsto \sigma(P_0, \ldots, P_{n-1}; x)$ passing over $\RR$. Obviously, it can only change at $x_0$ if one of the $P_i$ has a root at $x_0$\pause
\begin{itemize}
\item if $P_i \neq P_0$ has $x_0$ as a root, $P_{i-1}P_{i+1}(x_0) < 0$\\\pause
$\Rightarrow$ signs of $P_{i-1}$, $P_{i+1}$ are $\neq 0$, opposite, constant in NH of $x_0$\\\pause
$\Rightarrow$ signs $[1,\, \_\,, -1]$ or $[-1,\, \_\,, 1]$ in the entire NH, i.e. one sign change\\\pause
$\Rightarrow$ total number of sign changes not influenced\\[1mm]\pause
\item if $P_0$ has $x_0$ as root, $P_0P_1(x_0) < 0$ in left-NH of $x_0$, $>0$ in right-NH\\\pause
$\Rightarrow$ signs are different left of $x_0$ and the same right of $x_0$\\\pause
$\Rightarrow$ total number of sign changes decreases by one
\end{itemize}
\end{frame}

\begin{frame}{Proving Sturm's Theorem}
Formal proof: a lot of induction on the sequences and number of roots\\[2mm]
$\Longrightarrow$ messy and not terribly interesting, I'll spare you the details
\end{frame}

\begin{frame}{Proving Sturm's Theorem}
We now know that Sturm sequences can count roots.\\
But how do we construct one?
\end{frame}

@@

\begin{frame}{Construction Sturm sequences}
Why does it work? Nonobvious parts:\\[3mm]\pause

If $x_0$ is root of $P_0 = P$: $PP'(x_0) < 0$ in left-NH and $PP'(x_0) > 0$ in right-NH\\
-- pick neighbourhood without roots of $P_0$ and $P_1$ (except for $x_0$),\\
\hskip3mm apply mean value theorem\\[3mm]\pause

@@

\begin{frame}{Construction Sturm sequences}
This construction assumed no multiple roots.\\
What do we do if there are multiple roots?
\end{frame}

\begin{frame}{Construction Sturm sequences}
In case of multiple roots: Let $D := \mathrm{gcd}(P,P')$. Then:\pause
\begin{description}
\item[The obvious way:]\ \\
\begin{itemize}
\item compute canonical Sturm chain of $P / D$\\ (\enquote{divide out} multiple roots)
\end{itemize}\pause
\item[The clever way:]\ \\
\begin{itemize}
\item we can compute the canonical Sturm chain of $P$ and divide by $D$ afterwards\pause
\item but: if $D(x)\neq 0$, dividing by $D$ does not change the number of sign changes at $x$\pause
\item $\Longrightarrow$ unless the interval bounds are multiple roots, we can use the canonical construction without changes
\end{itemize}
\end{description}
\end{frame}

\begin{frame}{Making a decision procedure}
\textit{count\_roots\_between\ p\ a\ b}: picks the most efficient Sturm chain construction and:
$$\textit{count\_roots\_between}\ p\ a\ b\ \ =\ \ |\{x.\ a < x\ \wedge\ x \leq b\ \wedge\ p(x) = 0\}|$$\vskip2mm
\pause

@@

\begin{frame}{Making a decision procedure}
Some more fluff:
\begin{itemize}
\item and/or: count $x$ with $$P(x)=0\ \wedge\ Q(x)=0\hskip5mm\mathrm{or}\hskip5mm P(x)=0\ \vee\ Q(x)=0$$\pause
\item $\forall$-inequalities: $$\forall x.\ P(x) \neq Q(x)\ \wedge\ R(x) \neq S(x)\ \vee\ T(x)\neq U(x)$$\pause
\item $\forall$ with $<$ and $>$: $$\forall x.\ P(x) < Q(x)\ \wedge\ R(x) > S(x)\ \vee\ T(x)\neq U(x)$$
\end{itemize}
\end{frame}
